---
Name: Olivia
Portfolio: 10
Topic: Popdemo
Date: February 15, 2024
---

# Portfolio 10 - Popdemo


Popdemo - Presented by Steven
- Can project populations in two ways: deterministically and stochastically
- deterministic omits stochasity, super simple and not the most useful, so probably use stochastic projections
- Classify the matrix by either age-structured, or stage-structured
- Can directly adjust the probability matrix
- To list a species under the endangered species act, need to do some sort of future species projection, so these species are often the most well-understood, because you need to know what is happening to the population and why.
- A lot of time when we want to use these models, it is really hard to find the data to work with these models.
- If you get asked to do a Population Viability Analysis (PVA), most likely you will need to use matrix models to do so.



Portfolio Assignment:

```{r}
# package
library(popdemo)

# look at the polar bear data
Pbear
Pbear_vector <- c(0.106, 0.068, 0.106, 0.461, 0.151, 0.108) # from paper

# Task 1:
# Define a new matrix to feed into Aseq that takes the previous year’s sea ice
# status into account.
# Make good years more likely if the previous year was good (use 20% likelihood),
# and conversely # makes bad ice years more likely if the previous year was
# bad (use 80% likelihood).

p4 <- 0.2 # probability next year is a bad year given that this year was a good year - 20%
q4 <- 0.8 # probability next year is a bad year given that this year was a bad year - 80%
Pbear_task1_matrix <- matrix(c(rep(c((1-p4)/3, (1-p4)/3, (1-p4)/3, p4/2, p4/2), 3),
                      rep(c((1-q4)/3, (1-q4)/3, (1-q4)/3, q4/2, q4/2), 2)), 5, 5)

Pbear_task1_projection <- project(Pbear, Pbear_vector, Aseq = Pbear_task1_matrix, time = 50)
plot(Pbear_task1_projection)
```

## Answer
a) 20% probability: This defines that there is a 20% chance that next year is a bad year, given that this year was a good sea ice year.
b) 80% probability: This defines that there is an 80% chance that next year is a bad year, given that this year was a good sea ice year. 

When plotting results we see an overall downward trend, meaning that with these given constraints we see a decline in population size/density over time. The is a steep downward trend from ~ time 4 to time 15, where it then flattens out.


```{r}
# Task 2:
# Define 2 new parameters which describe the probability next year is a bad year
# given that this year was a good  (use 50% likelihood), and (b) describe the
# probability next year is a bad year given that this year was a bad year (use
# 80% likelihood again).

p5 <- 0.5 # probability next year is a bad year given that this year was a good year
q5 <- 0.8 # arbitrary object names
Pbear_task2_matrix <- matrix(c(rep(c((1-p5)/3, (1-p5)/3, (1-p5)/3, p5/2, p5/2), 3),
                      rep(c((1-q5)/3, (1-q5)/3, (1-q5)/3, q5/2, q5/2), 2)), 5, 5)

Pbear_task2_projection <- project(Pbear, Pbear_vector, Aseq = Pbear_task2_matrix, time = 50)
plot(Pbear_task2_projection)
```

## Answer
a) 50% probability: There is a 50% probability that next year is a bad year given that this year was a good year.
b) 80% probability: There is an 80% probability that next year is a bad year given that this year was a good year.


When plotting the results, we see an overall steep downward trend in population size/density over time. There is an even steeper initiial downward trend in this plot that in the previous one from time 0 to 15 because flattening out. Overall, not a good outlook for polarbears.







