---
Title: "Prortfolio 2"
Author: "Olivia Somhegyi"
Date: January 16, 2024
Topic: "Global Population Data"
---

# Portfolio 2 - Growth Population Data

## Growth curver package in R using world data

```{r}
# Exponential growth of Human Population
# Example
# Christopher L Jerde
# Notes:Data from https://ourworldindata.org/population-growth#introduction
########################################
```


```{r}
#Clear the R environment
rm(list = ls())

#Libraries needed
library(janitor) #cleans data and names
library(here) #allows for localized file directory
library(tidyverse) #makes R work nicely

#get the data
w_pop_data<-read_csv(here("data","population-and-demography.csv"))

w_pop_data_philippines<- w_pop_data |> clean_names() |>
  filter(country_name=="Philippines") |> # filter for Phillipines
  select(year, population) |> drop_na()

#clean the data for only the global human population
w_pop_data<- w_pop_data |> clean_names() |>
  filter(country_name=="Namibia") |> # filter for Namibia
  select(year, population) |> drop_na()
```


# Always plot your data
```{r}
#On the observed scale - Namibia
human_ts<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  xlim(1949, 2023)+
  ggtitle("Human Population in Namibia (1950-2022)")+
  theme_bw()
human_ts
```
Increasing - largely linear, with a few aspects that are exponential

```{r}
#On the observed scale - Philippines
human_ts_philippines<-ggplot(w_pop_data_philippines, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  xlim(1949, 2023)+
  ggtitle("Human Population in Philippines (1950-2022)")+
  theme_bw()
human_ts_philippines
```
The Philippines have logistic growth

```{r}
#on the transformed log scale - Namibia
human_ts_log<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  scale_y_continuous(trans="log")+ # log scale for linear transformation
  xlim(1949, 2023)+
  ggtitle("Human Population in Namibia (1950-2022)")+
  theme_bw()
human_ts_log
```
Increasing - largely linear, the slope decreases at the end of the time series indicating a drop off/slowing of population growth


# use lm() to find estimates
```{r}
human.lm_fit<- lm(log(population)~year,data=w_pop_data)
summary(human.lm_fit) #p-value < 2.2e-16; adjusted R-squared: 0.9889; results indicate model is a good fit.

NO<-exp(human.lm_fit$coefficients[1]) # calculate Nknot parameter (3.7e-16)
r<-human.lm_fit$coefficients[2] # calculate the growth rate coefficient (0.0249)
```


# Plot the data and the model
```{r}
# On the transformed scale - Namibia
human_ts_model_trans<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point()+
  xlab("Year")+
  ylab("Human count (N)")+
  scale_y_continuous(trans="log")+
  geom_smooth(method="lm",color="blue")+
  xlim(1949, 2023)+
  ggtitle("Human Population in Namibia (1950-2022)")+
  theme_bw()
human_ts_model_trans
```
Pretty good model fit, but the data is not quite linear. The model over estimate and under estimate often, but provides a general idea of the data trend even if it is not the best fit.



```{r}
# On the observed scale

# Predicted population values from the model
predicted_df <- data.frame(pop_pred = predict(human.lm_fit, w_pop_data), year=w_pop_data$year)
predicted_df<- predicted_df |> mutate(N_est=exp(pop_pred)) #transformed back to observed population estimates

# Namibia
human_ts_model<-ggplot(w_pop_data, aes(x=year, y=population))+
  geom_point(color="black")+
  geom_line(color='red',data = predicted_df, aes(x=year, y=N_est))+
  xlab("Year")+
  ylab("Human count (N)")+
  xlim(1949, 2023)+
  ggtitle("Human Population in Namibia (1950-2022)")+
  theme_bw()
human_ts_model
```
Good model fit until about year 1988, then the data deviates a bit from the model fit. Recent years (~2008-2022) aren't well fit for this model, but the first several years are very well fit.

