---
Name: Olivia
Portfolio: 3
Topic: Growth Rates
Date: January 30, 2024
---

# Growth Rates notes from class:

## lecture notes
growth rates over time
bacterial growth
value is population size

## you need to find r, K , and other parameters. Should always start at time = 0.
## function: 'fit_easylinear' finds the max growth rate on the exponential segment of the curve

Easy linear
## the whole plot does not need to be exponential, can be only a segment of the exponential curve and will take the moment in time with the highest growth rate
## coefficients being output: 'mumax' is r. r is the growth rate

Fit splines (non-parametric model)
## fit_spline - splines fit models piecewise to a dataset
## we hate splines
## this uses splines to find r
## the spline is essentially centered on a single datapoint


Fitting parametric nonlinear models
## logistic grow th rate model
p = needs a yknot (starting value), a mumax (r), and a K
TIME = time
y = population size

## To calculate p, it doesn't know where to start, so it needs some values to know where to start (kinda like nls)

Can apply this to multiple datasets
## spar argument is how smooth your splines will be
for logistic growth: use the 'all_growthmodels'
## plot dose reponse curves

Limitation
## Does not provide any info about model-best fit, so how would you choose which model?
## how to know which r value to start with? Hard to know, but it seems like in this particular case, a wide range of r will be okay so long as there aren't any local minmum or maximum.

## Assignment:
Find r and K for the bison dataset using at least three different models (Exponential, logistic, spline)
Try functions fit_easylinear, fit_growthmodels, and fit_Splines
Do the different models agree on the parameter values?


## Portfolio Assignment:

# load packages
```{r}
library(growthrates)
library(tidyverse)
library(ggplot2)
library(here)
library(janitor)
```

# read in the data
```{r}
bison <- read_csv(here("data", "bison.csv"))
```

# visualize data
```{r}
bison_gg <- ggplot(bison, aes(x = year, y = bison)) +
  geom_line() +
  geom_point() +
  labs(x = "Year",
       y = "Bison Count") +
  theme_minimal()

bison_gg
```

# applying fit_easylinear
```{r}
# use function 'fit_easylinear' 
fit_bison_linear <- fit_easylinear(bison$year, bison$bison)

# mumax is r
coef(fit_bison_linear)
#mumax = .1338

# plot results
plot(fit_bison_linear) # the plot will note which points have highest growth rate using the data
## plot fits data from 1970-1982 best.

# can also plot the log of results and compare between the two
```


# fit splines
```{r}
# use the funtion 'fit_spline'
fit_bison_spline <- fit_spline(bison$year, bison$bison)

# mumax is r
coef(fit_bison_spline)
# mumax = 0.5067

# plot the fitted spline
plot(fit_bison_spline)
## adds a spline connects each of the data points, first point at time 1970 is red
```


# fit growth models
```{r}
p <- c(y0 = 0.01, mumax = 0.15, K = 0.1)

fit_bison_growth <- fit_growthmodel(FUN = grow_logistic,
p = p,
bison$year,
bison$bison)

coef(fit_bison_growth)
# K = 2267.286

# visualize
plot(fit_bison_growth) # not the best fit
```


## each model type gets quite different parameter values, becuase it is fitting the data differently. To get a better idea of which function is best suited for the data, always perform an initial visual inspection to get an idea if growth is linear, exponential, etc.



